---
title: "Problem Set 5"
author: "Logan Fowler"
date: "2023-02-018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

# Disclaimer: this serves as a template for you to compose your answer for PS5. It is 100% hardcoded, so error may exists. Please refer to the official spec for a full list of questions and requirements.

# Instructions

This is a problem set about rmarkdown and plotting (using ggplot). Unlike the previous problem sets, this one does not give you a ready-made GH repo with a code file--it is now your task to create a repo and include your rmarkdown file in there.\

You should answer the questions below in that file, knit it, and submit both the compiled html and link to your repo on canvas.\

-   This problem sets asks you to write extensively when commenting your results. Please write clearly! Answer questions in a way that if the code chunks are hidden then the result is still readable!\
    -   All substantial questions need explanations. You do not have to explain the simple things like "how many rows are there in data", but if you make a plot of life expectancy, then you should explain what does the plot tell you.\
    -   Write explanations as markdown and use the styles like bold and italic as appropriate.\
-   Do not print too much results. It is all well to print a few lines of data for evaluation/demonstration purposes. But do not print dozens (or thousands!) of lines--no one bothers to look at that many numbers. You will lose points for annoying others (here your graders, but later potentially your boss).\
-   Do not make code lines too long. 80-100 characters is a good choice. Your grader may not be able to follow all the code if the line is too long--most of us are using small laptop screens! (And again--you want to keep your graders happy!)\

# Gapminder data

We use gapminder dataset, downloaded from [**https://www.gapminder.org/data/**](https://www.gapminder.org/data/), however, the data structure there is quire complex, please use the dataset provided on canvas (in files/data).\
The variables are:\
**name** country name\
**iso3** 3-letter country code\
**iso2** 2-letter country code\
**region** broad geographic region\
**sub-region** more precise region\
**intermediate-region**\
**time** year\
**totalPopulation** total population\
**GDP_PC** GDP per capita (constant 2010 US\$)\
**accessElectricity** Access to electricity (% of population)\
**agriculturalLand** Agricultural land (sq. km)\
**agricultureTractors** Agricultural machinery, tractors (count)\
**cerealProduction** Cereal production (metric tons)\
**feritilizerHa** Fertilizer consumption (kilograms per hectare of arable land)\
**fertilityRate** total fertility rate (births per woman)\
**lifeExpectancy** Life expectancy at birth, total (years)\
**childMortality** Mortality rate, under-5 (per 1,000 live births)\
**youthFemaleLiteracy** Literacy rate, youth female (% of females ages 15-24)\
**youthMaleLiteracy** Literacy rate, youth male (% of males ages 15-24)\
**adultLiteracy** Literacy rate, adult total (% of people ages 15 and above)\
**co2** CO2 emissions (kt)\
**greenhouseGases** Total greenhouse gas emissions (kt of CO2 equivalent)\
**co2_PC** CO2 emissions (metric tons per capita)\
**pm2.5_35** PM2.5 pollution, population exposed to levels exceeding WHO Interim Target-1 value 36ug/m3\
**battleDeaths** Battle-related deaths (number of people)\

# 1 Load and check data (5pt)

You first task is to do a very simple data check:

1.  (1pt) For solving the problems, and answering the questions, create a new rmarkdown document with an appropriate title. See [**https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating**](https://faculty.washington.edu/otoomet/info201-book/r-markdown.html#r-markdown-rstudio-creating).\

2.  (2pt) Load data. How many rows/columns do we have?\

```{r, message=FALSE}
library(tidyverse)
data <- read_delim("./gapminder.csv")
```

The data has `r nrow(data)` rows and `r ncol(data)` columns.

3.  (2pt) Print a small sample of data. Does it look OK?
```{r}
data %>% slice_sample(n = 5) %>% print()
```

The data looks to be okay, with consistent and reasonable values over multiple observations. There are some missing values in some observations, however.   

------------------------------------------------------------------------

# 2 Descriptive statistics (15pt)

1.  (3pt) How many countries are there in the dataset? Analyze all three: *iso3*, *iso2* and *name*.

There are `r data %>% distinct(iso3) %>% nrow()` *iso3* codes, `r data %>% distinct(iso2) %>% nrow()` *iso2* codes, and `r data %>% distinct(name) %>% filter(!is.na(name)) %>% nrow()` names present in the dataset.

2.  If you did this correctly, you saw that there are more iso-2 codes than names, and there are even more *iso3*-codes. What is going on? Can you find it out?

    (a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that correspond to more than one name? What are these countries?
    
    ```{r}
    data %>%
      group_by(iso2) %>%
      summarize(count = n_distinct(name)) %>%
      arrange(desc(count), .by_group = TRUE) %>%
      print()
    ```

    (b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have more than one iso3-code? What are these countries? Hint: two of these entitites are *CHANISL* and *NLD CURACAO*.
    
    ```{r}
    data %>%
      group_by(name) %>%
      summarize(count = n_distinct(iso3)) %>%
      arrange(desc(count), .by_group = TRUE) %>%
      head(5) %>%
      print()
    ```
    
    The issue seems to be that there are some nameless countries in the data that still possess iso2 and iso3 codes, resulting in there being more of those codes than names.

3.  (2pt) What is the minimum and maximum year in these data?

The minimum year in this data is `r min(data$time, na.rm = TRUE)`, and the maximum year is `r max(data$time, na.rm = TRUE)`.

------------------------------------------------------------------------

# 3 CO2 emissions (30pt)

Next, let's analyze CO2 emissions.

1.  (2pt) How many missing co2 emissions are there for each year? Analyze both missing *CO2* and *co2_PC*. Which years have most missing data?\

```{r}
data %>%
  group_by(time) %>%
  summarize(missing_co2 = sum(is.na(co2)),
            missing_co2_PC = sum(is.na(co2)),
            missing_total = missing_co2 +
                            missing_co2_PC) %>%
  arrange(desc(missing_total), .by_group = TRUE) %>%
  head(5) %>%
  print()
```

The years 2017, 2018, and 2019 have the most missing data on CO2.

2.  (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few more countries of your choice. Explain what do you see.\

```{r}
data %>%
  filter(iso3 == "USA" |
           iso3 == "CHN" |
           iso3 == "IND" |
           iso3 == "GRC" |
           iso3 == "FRA") %>%
  ggplot(aes(time, co2, col=name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions Per Capita", col = "Country Names")
```

We see that the CO2 emissions for China, the US, and India all rose over time, with China having the sharpest increase, India having the lowest, and the US being in the middle. France's emissions have actually fallen over time, and Greece's have remained relatively constant.

3.  (5pt) Now let's analyze the CO2 emissions per capita (*co2_PC*). Make a similar plot of the same countries. What does this figure suggest?\

```{r}
data %>%
  filter(iso3 == "USA" |
           iso3 == "CHN" |
           iso3 == "IND" |
           iso3 == "GRC" |
           iso3 == "FRA") %>%
  ggplot(aes(time, co2_PC, col=name)) +
  geom_line() +
  labs(x = "Year", y = "CO2 Emissions Per Capita", col = "Country Names")
```

This figure suggests that, when accounting for population, the US has much higher CO2 emissions per capita than other countries, and the emissions per capita for other countries tends to rise over time, except for in France, where it has been decreasing.

4.  (6pt) Compute average CO2 emissions per capita across the continents (assume *region* is the same as continent). Comment what do you see.\
    Note: just compute averages over countries and ignore the fact that countries are of different size.\
    Hint: Americas 2016 should be 4.80.\
    
```{r}
data.by_region <- data %>%
  group_by(region, time) %>%
  summarize(average_co2_PC = mean(co2_PC, na.rm = TRUE))
print(data.by_region)
```

The data seems to show that the average CO2 emissions per capita in each region/continent increases slightly over time.

5.  (7pt) Make a barplot where you show the previous results--average CO2 emissions per capita across continents in 1960 and 2016.\
    Hint: it should look something along these lines(**Please refer to the spec for the graph**)\
    
```{r}
data.by_region %>%
  filter(time == 1960 | time == 2016) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(region, average_co2_PC, fill=factor(time))) +
  geom_col(position = "dodge", na.rm = TRUE) +
  labs(x = "Continent", y = "Average CO2 Per Capita", fill = "Year")
```

6.  Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per capita) in 2019 for each continent? (Assume *region* is continent).\

```{r}
for (cur_region in unique(data$region)) {
  if (is.na(cur_region)) {
    next
  }
  print(cur_region)
  shmorb <- data %>%
    filter(time == 2016 &
             region == cur_region &
             !is.na(co2_PC)) %>%
    arrange(desc(co2_PC)) %>%
    select(name, co2_PC)
  shmorb %>% head(3) %>% print()
  shmorb %>% tail(3) %>% print()
}
```

------------------------------------------------------------------------

# 4 GDP per capita (50pt)

Let's look at GDP per capita (*GDP_PC*)

1.  (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for 1960. Make the point size dependent on the country size, and color those according to the continent. Feel free to adjust the plot in other ways to make it better.\
    Comment what do you see there.\
    
```{r}
data %>%
  filter(time == 1960 & !is.na(region)) %>%
  ggplot(aes(lifeExpectancy, GDP_PC, size = totalPopulation, col = region)) +
    geom_point() +
  labs(x = "Life Expectancy", y = "GDP Per Capita", size = "Population Size", col = "Geographic Region")
```

We see that life expectancy and GDP per capita are positively correlated, with one rising when the other rises.

2.  (4pt) Make a similar plot, but this time use 2019 data only.\

```{r}
data %>%
  filter(time == 2019 & !is.na(region)) %>%
  ggplot(aes(lifeExpectancy, GDP_PC, size = totalPopulation, col = region)) +
    geom_point() +
  labs(x = "Life Expectancy", y = "GDP Per Capita", size = "Population Size", col = "Geographic Region")
```

We see the same positive correlation again, except that the steepness of the correlation is stronger.

3.  (6pt) Compare these two plots and comment what do you see. How has world developed through the last 60 years?\

I see that the world has gotten a lot larger of a population, considering that the population size dots account for an order of magnitude more people. I also see that the overall life expectancy and overall GDP per capita have both improved greatly, as the dots are all shifted in the positive X and Y directions. Additionally, it seems that Asia has achieved more growth in life expectancy and GDP per capita compared to any other continent, as its dots have "shifted" in the positive direction the most.

4.  (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results fit with what do you see on the figures?\
    Note: here as *average*I mean just average over countries, ignore the fact that countries are of different size.\
    
```{r}
data %>%
  filter(time == 1960 | time == 2019) %>%
  filter(!is.na(region)) %>%
  group_by(region, time) %>%
  summarize(avg_le = mean(lifeExpectancy, na.rm = TRUE)) %>%
  print()
```

The results fit with what was seen in the figures, as across the board, the average life expectancy in each region increased by a substantial amount, as could be seen in the figures.

5.  (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results in the order of growth. Explain what do you see.\
    Hint: these data (data in long form) is not the simplest to compute growth. But you may want to check out the lag() function. And do not forget to group data by continent when using lag(), otherwise your results will be messed up! See [**https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute**](https://faculty.washington.edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute).\
    
```{r}
data %>%
  filter(time == 1960 | time == 2019) %>%
  filter(!is.na(region)) %>%
  group_by(region, time) %>%
  summarize(avg_le = mean(lifeExpectancy, na.rm = TRUE)) %>%
  mutate(diff = avg_le - lag(avg_le)) %>%
  filter(!is.na(diff)) %>%
  select(region, diff) %>%
  arrange(desc(diff)) %>%
  print()
```

As expected, every region had an increase in the average life expectancy over the 60 year period, with Asia having the most. However, Africa was really not far behind, which I did not expect from looking at the figures previously.

6.  (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both histograms on the same graph, see how well you can do it!\

```{r}
data %>%
  filter(time == 1960 | time == 2019) %>%
  ggplot(aes(GDP_PC, fill = factor(time))) +
  geom_histogram(position = "dodge", bins = 8) +
  labs(x = "GDP Per Capita", y = "Count", fill = "Year")
```

7.  (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When counting from top.)\
    Hint: check out the function rank()!\
    Hint2: 17 for 1960.\
    
```{r}
data %>%
  filter(time == 1960 | time == 2019) %>%
  filter(!is.na(name)) %>%
  group_by(time) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  filter(iso3 == "USA") %>%
  select(name, time, lifeExpectancy, rank) %>%
  print()
```

8.  (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a bit. But we also have more countries in 2019--what about the relative rank divided by the corresponding number of countries that have LE data in the corresponding year?\
    Hint: 0.0904 for 1960.\
    
```{r}
data %>%
  filter(time == 1960 | time == 2019) %>%
  filter(!is.na(name) & !is.na(lifeExpectancy)) %>%
  group_by(time) %>%
  mutate(rank = rank(desc(lifeExpectancy))) %>%
  mutate(ratio = rank / n()) %>%
  filter(iso3 == "USA") %>%
  select(name, time, rank, ratio) %>%
  print()
```

------------------------------------------------------------------------

Finally tell us how many hours did you spend on this PS.

I spent around 4 hours on this PS.
